The demographic costs of sexually antagonistic selection in partially selfing populations

### 2022-01-05 doc/ResponseLetter-R1.tex

\section*{}
\noindent To the Ed., AE, \& Editorial Board
\bigskip

\noindent Please consider this revision of our manuscript (MS \#60734), titled "The demographic costs of sexually antagonistic selection in partially selfing populations" for publication in \textit{The American Naturalist}. 
\bigskip

\noindent First of all, we want to thank the Ed., AE, and EiC for their thoughtful and forthright responses regarding the editorial delay on our manuscript. We completely understand that pandemic-related personal issues have been and continue to be extremely disruptive in these difficult times, and that editorial duties for our journals is a volunteer service. To our AE, Dr.~Orive, we sincerely hope that you and your family are well, and we truly appreciate the time and effort you took to evaluate our paper when you clearly had more imporant things to deal with.
\bigskip

\noindent We received very helpful and constructive comments on our first submission and have now made a variety of major revisions to the MS in order to address them. First, we have made significant changes to our introduction to develop conceptual links between our model and previous work on evolutionarily explicit structured models and empirical research on sexually antagonistic selection. Second, we have tried to lessen the burden on non-theoretical readers by shifting development of the general model to the Supplementary Material, by following R2's many helpful suggestions regarding the presentation of the 2-stage model, and by including a new explanatory figure to help clarify the genetic and demographic outcomes from our model. Third, we have revised several figures to show population intrinsic growth rates ($\lambda$) rather than extinction boundaries. Finally, we have addressed the impacts of density dependence by ({\itshape i}) developing a new version of our model which includes simple density-dependence and reanalyzing the {\itshape M. guttatus} example; and ({\itshape ii}) expanding our discussion of density-dependence to address these additions. Overall, we feel that these various revisions have greatly improved the paper. 
\bigskip

@@

\begin{quote}
	{\itshape This is a very good and totally valid point. Honestly, we believe a separate paper is necessary to deal with the many consequences of density dependence for our model predictions. However, we also agree that sidestepping this issue completely isn't good enough. We have now made two main changes to address density dependence. First, have added a new {\bf Appendix B} where we briefly show how density dependence can be incorporated into the model, and reanalyze our {\itshape M. guttatus} example with a simple density-dependent version of the model. Second, we have expanded our methods, results, and especially the discussion to more fully address the implications of density-dependence, with reference to our modified figures and the new {\bf Appendix B} (see {\bf L.128--134; 456--459; 529--561}).}
\end{quote}

@@

Related to that, and without doing more modeling, one quantity of interest you can present is the actual growth rate, and not just the region of viability/inviability. In a world where we acknowledge density dependence as a possibility, how far above or below one growth rate is can be informative about whether you actually expect extinction or depressed densities.

\begin{quote}
	{\itshape This is great suggestion which inspired us to make several major changes. We now include a new illustrative figure ({\bf fig.~1}), and revised versions of figs.~1 \&~4 (now {\bf figs.~S1 \& 5}) which show heatmaps of absolute values for predicted $\lambda$ instead of just extinction boundaries. Unfortunately, when presenting $\lambda$ in this way, it is no longer possible to show multiple values of maximum fertility ($f$) on a single panel. In particular, our old fig.~1 now requires 18 panels to illustrate $\lambda$ for the same set of parameter combinations. We felt that this resulted in an unnecessarily large and complicated figure. For simplicity, we have decided to retain our original version of fig.1 (now {\bf fig. 2}) in the main text and include the modified version showing 18 heatmaps for $\lambda$ as a supplementary figure (now {\bf fig.~S1}). We hope that, with the new illustrative {\bf fig.~1}, the extinction boundaries presented in {\bf fig.~2} will be easier for readers to understand. However, if the editors and reviewers prefer the full heatmaps, we will happily swap {\bf fig.~2} with {\bf fig.~S1}. Lastly, we agree that the {\itshape M. guttatus} example is greatly improved by showing $\lambda$, and we now present the modified version of fig. 4 (now {\bf fig. 5}) in the main text}.
\end{quote}

@@

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\section*{Responses to Associate Editor Maria Orive}

Both reviewers found the work interesting and potentially to be of strong interest to the readership of AmNat, and I agree. However, Reviewer 1 brings up an important point regarding the framing of the work and placing it within a broader context of the literature on evolutionarily explicit structured models, which I think is important for the authors to do in order to make the paper accessible to and of interest for the overall readership of AmNat. There are other points that the reviewers bring up that are important to consider, but I think this one needs to be addressed carefully.

\begin{quote}
	{\itshape We can see now that in our efforts to appeal to both demographers and evolutionary biologists studying sexual conflict, we insufficiently discussed the relevant literature on both counts. We have restructured our introduction to better engage with these fields without lengthening it (see, in particular, {\bf L.43--51, L.58--67}). See also our specific responses to Reviewer 1. }
\end{quote}

@@

Reviewer 2 has some very nice suggestions regarding ideas for simplifying and clarifying the model description.

\begin{quote}
	{\itshape Agreed. Please see our specific responses to Reviewer 2 below.}
\end{quote}

In addition to the clarifications/edits and other points raised by the reviewers, I include a handful of minor edits/typos here of my own:
\bigskip

\noindent (1) Abstract, perhaps add a very general and short explanation of the term sexually antagonistic polymorphism.
\begin{quote}
	{\itshape We have added a brief explanation of sexually antagonistic selection to the beginning of the Abstract, which we felt was lacking in the first draft, and also allows us to more clearly link SA polymorphism with its' demographic consequences ({\bf L.2--6}).}
\end{quote}

@@

\begin{quote}
	{\itshape Thank you again for your efforts and consideration in spite of these challenges. We really appreciate it.}
\end{quote}

\noindent xxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxxx

\subsection*{Reviewer \#1:}

The manuscript reports the derivation, analysis, and example application of an extended frequency-dependent matrix model to explore links between sexual antagonistic selection, population persistence, and the maintenance of heritable variation. The paper is very well written and is straightforward to follow. The derivations are correct, although I did not try to derive all the equations in the appendices. My one comment is how might environmental stochasticity in the rate of selfing and outcrossing impact general conclusions? 
\begin{quote}
	{\itshape We now briefly discuss this possibility in the Discussion ({\bf L.581--587}).}
\end{quote}

@@

The manuscript adds to a growing and important literature where eco-evolutionary feedbacks are examined in structured models, some of which incorporate evolution explicitly. It also contributes nicely to a body of empirical and theoretical work on sexually antagonistic selection.

\begin{quote}
	{\itshape Thank you for these encouraging words.}
\end{quote}

As written the manuscript will not appeal to the broad readership of the American Naturalist because the weakness of the manuscript is the author makes little attempt to fit the work into the broader relevant literature. Instead, the paper has been written for a small audience of experts interested in the demographic consequences of sexually antagonistic selection. As written, the work is suited to a journal such as Journal of Theoretical Biology. This is disappointing, and the authors do themselves no favors by writing for such a niche audience. The manuscript could be well-suited to the American Naturalist if the authors were to engage with the broader literature. This can be done by extending the introduction to link this work to the wider literature on evolutionarily explicit structured models, and to the large empirical literature on sexually antagonistic selection. If they are able to do this appropriately, I believe this paper would make a strong contribution to the American Naturalist.

\begin{quote}
	{\itshape This is a very important comment and we thank you for pointing out this major shortcoming. Our goal when crafting the first draft was to find a middle ground that would appeal to both demographers and people studying sexual conflict. It is clear that we didn't quite achieve that goal. Rather, by making compromises in the introduction we ended up lacking scholarship on both sides. To address your main concern, we have restructured the Introduction to more directly connect with the broader literature on evolutionarily explicit structured models (see {\bf L.58--67}) and now also provide a stronger link with the empirical sexually antagonistic selection literature (see {\bf L.43--51}). We hope these changes, along with a more streamlined opening ({\bf L.20--23}) and presentation of SA selection ({\bf L.34--54}) succeed in engaging with the broader AmNat readership.}
\end{quote}

@@

\begin{quote}
	{\itshape Thank you for the very encouraging assessment.}
\end{quote}

\noindent I have one main comment:

\noindent-- Discussion: Conclusions are drawn before limitations of the model are mentioned. However, it seems to me that it would be very important to discuss the effect of those limitations on the results themselves and therefore discuss the generality of the results. For instance, what is the potential effect of female dominance in the marriage function on the results and in particular on conclusion of lines 477-480.

\begin{quote}
	{\itshape While we feel that there is strong precedence for discussing the key results of a study before diving into the various caveats for with the analysis, your point is well taken that we need to more clearly acknowledge these limitations in the Methods section and direct readers to where they are discussed. In our Methods section, we are now careful to point readers to the various Appendixes, Supplements, and places in the Discussion where these limitations are dealt with (e.g., see {\bf L.94--98, 103--109, 128--134, 183--186, 187--189, 196--197, 207--212, 274--275}), and we have added more explicit discussion of several of our key assumptions in the discussion ({\bf e.g., L.519--581}). We discuss the causes and potential effects of incorporating male-limitation of reproductive success on {\bf L.550--561}. We are reluctant to speculate much further on the potential effects of incorporating male gamete production because this will simultaneously influence both the invasion conditions and population growth rates in different regions of parameter space}.
\end{quote}

@@

\end{itemize}

\noindent -- Line 19: I find it weird to write that the "process of adaptation can be impeded by ...changing environmental conditions".

\begin{itemize}
	\item One organism can be adapted to an environment (as in a series of environmental parameters) but also to changes in that environment (plasticity, genetic variance etc ...). It is actually from this consideration that a lot of studies on adaptation and adaptive dynamics.
	\item I think this sentence should be rewritten to make it clear that it is not changes in environmental conditions per se that impede the adaptive process but some of their characteristics (speed, range etc.)
\end{itemize}

\begin{quote}
	{\itshape These are good points, and that was awkward wording on our part. While revising the introduction, we felt that much of this first paragraph was tangential to the key concepts we wanted to introduce. We have now re-written and shortened it to more quickly and concisely introduce several concepts which we feel are more relevant to the rest of the introduction ({\bf L.20--24}). The offending sentence has been removed.}
\end{quote}

@@

\noindent -- Line 23: The link between genetic constraints and genetic trade-offs should be clarified. Or rather a definition of each term provided, for the reader to understand this sentence. I suspect that, for many a reader of AmNat, genetic constraints and genetic trade-offs are equivalent concepts.

\begin{quote}
	{\itshape Agreed. We have rephrased to avoid the term "trade-off" which tends to over-simplify the concept in our opinion ({\bf L.23--24}).}
\end{quote}

\noindent -- Line 30: "fitness trade-offs ... provide ...mechanism for the maintenance of genetic variation". This requires a few words of explanation. Is that a general statement about (fitness) trade-offs? or is it actually particular to SA trade-offs or to trade-offs between fixed heterogeneity components?
\begin{itemize}
	\item For instance, a fitness trade off between fertility at 3 years and fertility at 6 years will have no positive effect on genetic variance. To the contrary, it may lead the organism to evolve an optimal strategy accounting for overall fitness and the trade-offs, vis-à-vis investments in reproduction at those different stages in the life cycle. And as such this may lead to the selection of individuals with such a strategy and therefore to low genetic variance.
\end{itemize}

\begin{quote}
	{\itshape We have combined this paragraph with the previous one in this revision. We have edited the opening of this combined paragraph to be more specific ({\bf L.34--39}), which we hope also addresses this comment.}
\end{quote}

@@

\noindent -- Line 264: this is an important point that is a bit unclear here, according to me. Could you clarify that what you call demographic rates are the transitions in the overall model and so include both demography and genetics? Clarify also that the model is not density dependent because it only depends on frequencies ... and maybe relate it to a general equation where A does not depend on n per se but on the q frequencies.

\begin{quote}
	{\itshape We have added a few lines to discuss frequency vs density dependence shortly after introducing the population projection matrix in Eq(3). Additionally, we have tried to improve this section by discussing that the survival and fertility rates ("demographic rates") are density-indepedent, and that the allele frequency dynamics make the model frequency dependent ({\bf L.128--134}).}
\end{quote}

@@

- Line 35: SA selection arises also from non-genetic trade-offs, which makes the formulation of that sentence a bit weird.

\begin{quote}
	{\itshape See comment above regarding {\bf L.34--39}.}
\end{quote}

### 2022-01-05 doc/SA-Hermaphrodites-wDemography-R1.tex
Final check before resubmitting
update expired anonymous GitHub link

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. \textcolor{blue}{We focus our presentation on a two-stage example of the model and discuss its' underlying biological assumptions. The derivation of the general model for any number of age or stage classes is provided in the Online Supplementary Material.} The derivation follows closely the logic presented in \citet{deVriesCaswell2019a, deVriesCaswell2019b}, and in fact, it reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing \textcolor{blue}{(when selfing is not allowed)}. All computer code necessary to reproduce the results are available at \url{https://anonymous.4open.science/r/SA-Hermaphrodites-wDemography-BF66}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.

>>>>

Here, we briefly describe a matrix model incorporating multiple life-cycle stages and a single diallelic locus under SA selection on male and female fertility for populations of partially-selfing simultaneous hermaphrodites. \textcolor{blue}{We focus our presentation on a two-stage example of the model and discuss its' underlying biological assumptions. The derivation of the general model for any number of age or stage classes is provided in the Online Supplementary Material.} The derivation follows closely the logic presented in \citet{deVriesCaswell2019a, deVriesCaswell2019b}, and in fact, it reduces to the two-sex stage-structured model of \citet{deVriesCaswell2019a} under obligate outcrossing \textcolor{blue}{(when selfing is not allowed)}. All computer code necessary to reproduce the results are available at \url{https://anonymous.4open.science/r/SA-Hermaphrodites-wDemography-4FFF}.%\url{https://github.com/colin-olito/SA-Hermaphrodites-wDemography}.
